Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-03-09 demystify.tex
updates

The most concrete evidence of research ability is having \textbf{papers in reputable international journals or conferences}.
Having published papers, especially at top venues, is a sign that the applicant has successfully involved in research.  Students  accepted to top schools often have multiple first-authored papers at top venues.

@@

Most PhD programs will require at least \textbf{two LORs}. Having a strong letter from an internationally recognized researcher will greatly strengthen your application. In fact, a strong letter from a well-known researcher can outweigh the lack of publications.

However, obtaining such letters
can be challenging for international students, who might not have much interactions with such experts. So it is fine to have a letter from people (e.g., profs, researchers, postdocs who mentored you) that know you well enough to talk about \emph{your research experience and capabilities}. Many students get letters from supervisors from companies where they did internships or are working at. It is OK as long as it is a research-based personalized letter (once again, we are talking about PhD applications, not MS). Again, the emphasis here is \emph{research}, i.e., the letters should describe your research experiences and potentials. Letters focusing your course work or non-research projects at a company won't carry much weight.

@@

\paragraph{Generic Letters} When the writers do not know much about the applicants (e.g., just taking some course with them or not making any impression to write about), they might write a generic letter, which is not useful and also considered weak.  It might be a good idea to directly ask if the prof. is willing to write a \emph{strong} letter for you. If not, then you should ask someone else.  For example, if a student I don't know well ask me to write letter for them, I will explicitly tell them I don't know them that well to write much about them, and such a short, generic, and weak letter will not help their case.

@@

\begin{commentbox}[Didier]
  \emph{Should letter writers have PhDs?}  In Rwanda, a lot of students interact more with teaching faculty who might not have PhD.
  \tcblower
  \textbf{Vu}: This is an interesting and useful detail that US faculty might not be aware of and students should mention about this in their statements. In general, I think it should be fine as long as that person can properly evaluate your research ability.
\end{commentbox}

\paragraph{Reminding Writers} After you submit your application, you should tell your writers about that and remind them to submit their letters if they haven't done so. But don't send too many reminders as that can be annoying to the writers.

@@

\paragraph{VEF 2.0} For Vietnamese students, it's worth mentioning about the \href{https://vef2.org/}{VEF2.0 program}, which has helped many good students in gaining admission to top CS PhD programs. VEF2.0 follows an interesting model where US faculty members from leading institutions are invited to conduct rigorous interviews with VEF students and subsequently provide reference letters on their behalf. Despite the limited interaction between the interviewer and the interviewee (primarily confined to the interview itself), these reference letters are generally effective as they have helped many students getting into top PhD programs in the US.

@@

While you might not have control over LORs or \hyperref[sec:your-school]{where your go to school}, you do over your
statement of purpose (SOP) or personal statement\footnote{Some schools separate these documents and ask you to write both: SOP, which focuses on research experiences, and Personal Statement, which is everything more personal, e.g., why PhD, challenges, diversity, etc}! So write it well because it could make a big difference.
In your statement, you have the opportunity to make your application stand out and unique, which can make you \emph{fit} the CS program you're applying to, even if you don't have very strong research experience.
A well-written SOP also shows that you can communicate, which is very important in research, and that you can effectively teach and communicate with students, which is important for TA funding (see \autoref{sec:funding}).
% is important because if you need (GTA) funding, it will provide evidence
% that you can teach and communicate with students.

@@

You should also talk about things that adcom members might not know about and can help make you different than the other 1000 applicants, e.g.,  your personal Github project with 1K+ stars or your regular contributions to well-known open-source projects (see \autoref{sec:improve-your-chance} for increasing your admission chance).

@@

Typically, an interview takes about 15--30 minutes, and one important aspect of evaluation is your ability to effectively communicate, including speaking and understanding English. A professor might want to know more about your research background or even ask you to read a paper (so be sure to do it!). During the interview, you will have the opportunity to ask questions about the university. It's essential to prepare some thoughtful questions, similar to a job interview.

@@

Some schools send out admission letters in batches, some do rolling admission, and some do not send anything out (e.g., you're on their waitlist, or they might not send rejection letters).  However, in general, you should hear back from most schools by around March.

@@

By around March you should hear back from most PhD programs you applied to. If you haven't heard back, reach out via email and ask about the status of your application.
If you receive offers, congratulations!  Now you're at a different game because the schools that have admitted you will try to get you to accept them!  Important factors to consider include the reputation of schools and professors (\autoref{sec:schoolsandprofs}), and funding availability (\autoref{sec:funding}). You will have to make your decision by around \emph{April 15} (\href{https://cgsnet.org/wp-content/uploads/2024/01/CGS_April15_Resolution_Jan312024.pdf}{why this date?}).

@@

You can learn about CS programs and research expertise of faculty using resources such as \href{https://csrankings.org}{CSRankings.org}. You will be very surprised to learn that a school that you didn't know much about can have very strong research in your interested topic (and vice versa, a school you thought highly about might have no faculty working in the research field you're interested in). This is also a good way to learn about individual faculty (who works on what) and well-known CS conferences\footnote{In CS (and probably only in CS), conferences, not journals, are often the main venue to publish research finding.}. \autoref{sec:ranking} gives the top 50 CS programs in the US according to CSRankings.

@@

However, in general, rankings can be superficial and you need to do more research to be informed and make better decision. For example, if you get admissions to several places, you should consider attending Open Houses (\autoref{sec:accepted}) and contact profs. that you're interested in at those place and talk to them. They would be more willing to chat with you now that you have been admitted.  Ask them questions about their work, how they manage students, their expectations. You can even ask to contact their students. See more in \autoref{sec:accepted} on what to do after getting admitted.

@@

\emph{Course work}: in the US you will spend the first couple of years taking classes and explore potential adviser and research topics. In other countries, you (who already have an MS) start your research right away, e.g., you immediately work on the research project you proposed with the adviser you chose. Also, in the US you also have to pass a series of "exams", e.g., qualifying exam, comprehensive exam, thesis proposal defense\footnote{The word "ABD" (all but dissertation) is used in the US to refer to a PhD candidate who have finished all course work and exams and only need to write and defend their dissertation.}. In other countries, you do not have to do these exams or only do a few of them.

\emph{Funding}:  In many countries stipend comes from the university or from the gov't. These funding often have a fixed duration (e.g., 3 years).  In the US, stipend (e.g., RA) comes directly from your adviser (no fixed duration).  There are also fewer TA opportunities in the European universities compared to the US.

@@

\emph{Work-life balance}: PhD students in the US are often said to be overworked compared to other countries, e.g., in Europe.  This is partly due to the longer PhD program and that US PhD students are often paid through TA, which requires them to do TA in addition to their own research. In contrast, PhD students in other countries are often paid through fellowships, which do not require them to do TA.

>>>>

A concrete evidence of research ability is having \textbf{papers in reputable international journals or conferences}.
Having published papers, especially at top venues, is a sign that the applicant has successfully involved in research.  Students  accepted to top schools often have multiple first-authored papers at top venues.

@@

Most PhD programs will require at least \textbf{two LORs}. Having a strong letter from an internationally recognized researcher will \emph{greatly strengthen} your application. In fact, a strong letter from a well-known researcher can outweigh the lack of publications (or low GPA or unknown school).

However, obtaining letters from famous people 
can be challenging for international students, who might not have much interactions with such experts. So it is fine to have a letter from people (e.g., profs, researchers, postdocs who mentored you) that know you well enough to talk about \emph{your research experience and capabilities}. Many students get letters from supervisors from companies where they did internships or are working at. It is OK as long as it is a research-based personalized letter (once again, we are talking about PhD applications, not MS). Again, the emphasis here is \emph{research}, i.e., the letters should describe your research experiences and potentials. Letters focusing your course work or non-research projects at a company won't carry much weight.

@@

\paragraph{Generic Letters} When the writers do not know much about the applicants (e.g., just taking some course with them or not making any impression to write about), they might write a generic and short letter, which is not useful and also considered weak.  It might be a good idea to directly ask if the prof. is willing to write a \emph{strong} letter for you. If not, then you should ask someone else.  For example, if a student I don't know well ask me to write letter for them, I will explicitly tell them I don't know them that well to write much about them, and such a short, generic, and weak letter will not help their case.

@@

\begin{commentbox}[Didier]
  \emph{Should letter writers have PhDs?}  In Rwanda, a lot of students interact more with teaching faculty who might not have PhD.
  \tcblower
  \textbf{Vu}: This is an interesting and useful detail that US faculty might not be aware of and students should mention about this in their statements. In general, someone with a PhD has been through the research process and therefore can better evaluate your research ability.  But if you do not have such people, then someone who can properly evaluate your research ability is still OK.
\end{commentbox}

\paragraph{Reminding Writers} After you submit your application, you should tell your writers about that and let them know they will soon receive an email from the university to submit their letters.  You should also remind them about the deadline and ask them to submit their letters on time.  But do not send too many reminders as that can be annoying to the writers.

@@

\paragraph{VEF 2.0} For Vietnamese students, it's worth mentioning about the \href{https://vef2.org/}{VEF2.0 program}, which has helped many good students in gaining admission to top CS PhD programs. VEF2.0 follows an interesting model where US faculty members from leading institutions are invited to conduct rigorous interviews with VEF students and subsequently provide reference letters on their behalf. Despite the limited interaction between the interviewer and the interviewee (primarily confined to the interview itself), these reference letters are generally effective and have helped many students getting into top PhD programs in the US.

@@

While you might not have control over LORs (\autoref{sec:lor}) or where your go to school (\autoref{sec:your-school}), you do over your
statement of purpose (SOP) or personal statement\footnote{Some schools separate these documents and ask you to write both: SOP, which focuses on research experiences, and Personal Statement, which is everything more personal, e.g., why PhD, challenges, diversity, etc}! So write it well because it could make a big difference.
In your statement, you have the opportunity to make your application stand out and unique, which can make you \emph{fit} the CS program you're applying to, even if you don't have very strong research experience.
A well-written SOP also shows that you can communicate, which is very important in research, and that you can effectively teach and communicate with students, which is important for TA funding (see \autoref{sec:funding}).
% is important because if you need (GTA) funding, it will provide evidence
% that you can teach and communicate with students.

@@

You should also talk about things that adcom members might not know about and can help make you different than the other 1000 applicants, e.g., you are member of a minority or LBGTQ+ group (\autoref{sec:urm}), your personal Github project with 1K+ stars or your regular contributions to well-known open-source projects (see \autoref{sec:improve-your-chance} for increasing your admission chance).

@@

Typically, an interview takes about 15--30 minutes, and one important aspect of evaluation is your ability to effectively communicate, including speaking and understanding English. 
In general treat it as an informal chat, but be prepared to talk about your research experiences and interests (so do have \emph{elevator pitch} about your experiences and research). 
A professor might want to know more about your research background or even ask you to read a paper (so be sure to do it!). During the interview, you will have the opportunity to ask questions about the university. It's essential to prepare some thoughtful questions, similar to a job interview.

@@

Some schools send out admission letters in batches, some do rolling admission, and some do not send anything out (e.g., you're on their waitlist, or they might not send rejection letters).  However, in general, you should hear back from most schools by around March (but some schools might take longer!).

@@

By around March you should hear back from most PhD programs you applied to. If you haven't heard back, reach out via email and ask about the status of your application.
If you receive offers, congratulations!  Now you're at a different game because the schools that have admitted you will try to get you to accept them!  Important factors to consider include the reputation of schools and professors (\autoref{sec:schoolsandprofs}), and funding availability (\autoref{sec:funding}). You will likely have to make your decision by around \emph{April 15} (\href{https://cgsnet.org/wp-content/uploads/2024/01/CGS_April15_Resolution_Jan312024.pdf}{why this date?}).

@@

You can learn about CS programs and research expertise of faculty using resources such as \href{https://csrankings.org}{CSRankings.org}. You will be very surprised to learn that a school that you didn't know much about can have very strong research in your interested topic (and vice versa, a school you thought highly about might have no faculty working in the research field you're interested in). This is also a good way to learn about individual faculty (who works on what) and well-known CS conferences\footnote{In CS (and probably only in CS), conferences, not journals, are often the main venue to publish research finding.}. %\autoref{sec:ranking} gives the top 50 CS programs in the US according to CSRankings.

@@

While I find CSrankings useful, it is not perfect, and in general, rankings are superficial and you need to do more research to be informed and make better decision. For example, if you get admissions to several places, you should consider attending Open Houses (\autoref{sec:accepted}) and contact profs. that you're interested in at those place and talk to them. They would be more willing to chat with you now that you have been admitted.  Ask them questions about their work, how they manage students, their expectations. You can even ask to contact their students. See more in \autoref{sec:accepted} on what to do after getting admitted.

@@

\emph{Course work}: in the US you will spend the first couple of years taking classes and explore potential adviser and research topics. In other countries, you (who already have an MS) start your research right away, e.g., you immediately work on the research project you proposed with the adviser you chose. Also, in the US you also have to pass a series of "exams", e.g., qualifying exam, comprehensive exam, thesis proposal defense\footnote{The word "ABD" (all but dissertation) is used in the US to refer to a PhD candidate who have finished all course work and exams and only need to write and defend their dissertation.}. In other countries, you might not have these exams or only have to do a few of them.

\emph{Funding}:  In many countries stipend comes from the university or from the gov't. These funding often have a fixed duration (e.g., 3 or 4 years).  In the US (\autoref{sec:funding}), stipend (e.g., RA) comes directly from your adviser (no fixed duration).  There are also fewer TA opportunities in the European universities compared to the US.

@@

\emph{Work-life balance}: PhD students in the US are often said to be overworked compared to other countries, e.g., in Europe.  This is partly due to the longer PhD program and that US PhD students are often paid through TA, which requires them to do TA in addition to their own research. In contrast, PhD students in other countries are often paid through fellowships, which might not require doing TA.
